---
title: "Speeding up R for Calculating Euclidean Distances"
author: 'Gibril'
output: 
    html_document:
        toc : TRUE
        toc_float: TRUE

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The task is to calculate euclidean distances between rows in a matrix. It is divided into two subproblems, one with groups and without groups. The following `R` packages have been used:

* `Rcpp`
* `microbenchmark`: for timings

A base `R` function was used as a benchmark for the timings. Three `C` source codes were used which were sourced into `R` via it's `C` API and a `C++` script via `Rcpp` library. 

Function  | Language
----------|---------
naive_dist| `R`
dot_dist  | `R`
fastdist2 | `C++`
eucDist   | `C` & `R`
r_euc_dist| `C` & `R`
cDist     | `C` & `R`


# Timings : Ungrouped 
Below are sample outputs from the functions for a matrix with 2000 entries for the ungrouped problem and relative timings of each of the functions for different matrix sizes. All timings were in milliseconds.

```{r echo=FALSE, warning=FALSE, tidy=TRUE, fig.width=9, comment=''}

library(microbenchmark)
library(Rcpp)

set.seed(1)

sourceCpp('~/Desktop/2019/c_c++/abdul_code/fastdist.cpp')

dyn.load('~/Desktop/2019/c_c++/euc_dist.so')
dyn.load('~/Desktop/2019/c_c++/abdul_code/eucDist.so')
dyn.load('~/Desktop/2019/c_c++/cDist.so')


source('~/Desktop/2019/r_scripts/functions/r_euc_dist.R')
source('~/Desktop/2019/r_scripts/functions/dot_dist.R')
source('~/Desktop/2019/r_scripts/functions/naive_dist.R')
source('~/Desktop/2019/r_scripts/functions/eucDist.R')
source('~/Desktop/2019/r_scripts/functions/cDist.R')



M = matrix(rnorm(2000, mean = 10, sd = 10), ncol = 2)

microbenchmark(naive_dist(M), 
               dist(M), 
               r_euc_dist(M),
               dot_dist(M),
               fastdist2(M),
               eucDist(M),
               cDist(M),
               unit = 'ms')

print('naive_dist')
head(naive_dist(M))
print('dist')
head(dist(M))
print('r_euc_dist')
head(r_euc_dist(M))
print('dot_dist')
head(dot_dist(M))
print('fastdist2')
head(fastdist2(M))
print('eucDist')
head(eucDist(M))
print('cDist')
head(cDist(M))
              

# timing data
pp = list(m10 = 0.0146535 / c(0.0450285, 0.0409605, 0.0146535, 0.0098735, 0.0109350, 0.0028890, 0.0080135),
          m100 = 0.0537830/ c(4.1130170, 3.5141160, 0.0537830, 0.0242315, 0.0262445, 0.0450945, 0.0165050),
          m1000 = 2.8566960/ c(408.1175120, 348.2601585, 2.8566960, 1.0159860, 0.7056655, 4.0971850, 0.4011935),
          m10000 = 337.96846/ c(41327.70777, 35103.25143, 337.96846, 153.70588, 121.82413, 457.65282,88.63522))




ppdat = as.data.frame(pp)

xcord = c(5,50,500, 5000)
plot(xcord, 
     ppdat[1,], 
     ylim = c(0,10), 
     type = 'l', 
     col = 'red',
     lwd = 3.5,
     xlab = 'No. of rows in Matrix',
     ylab = 'Relative time')


points(xcord, ppdat[2,], type = 'l', col = 'cyan', lwd = 2)
points(xcord, ppdat[3,], type = 'l', col = 'green', lwd = 2, lty = 2)
points(xcord, ppdat[4,], type = 'l', col = 'gray', lwd = 2)
points(xcord, ppdat[5,], type = 'l', col = 'brown', lwd = 2)
points(xcord, ppdat[6,], type = 'l', col = 'violet', lwd = 2)
points(xcord, ppdat[7,], type = 'l', col = 'black', lwd = 2)

legend('topright', 
       title = 'Functions', 
       legend = c('naive_dist', 'dot_dist', 'dist',
                  'eucDist', 'r_euc_dist', 'fastdist2', 'cDist'),
       lwd = 3, 
       col = c('red', 'cyan', 'green','gray', 'brown', 'violet', 'black'),
       cex = 0.8)


```

# Timings: Grouped

## Output
Below are sample outputs from the functions `dist`, `cDist`, `naive_dist`, `r_euc_dist` and `eucDist` for the grouped problem. The outputs from all the functions are the same so only one is shown. The functions `dot_dist` and `fastdist2` could not be modified to allow for a vectorised calculation of distances within each group.
```{r grouped, echo=F, comment=''}
 grp = sample(1:50, 1000, replace = TRUE)
    M   = matrix(rnorm(2000, mean = 10, sd = 10), ncol = 2)
    M   = cbind(M, grp)
    
    by(M, grp, cDist)
    
    
```

## Relative times
The graph below shows the relative timings of each of the functions for different matrix sizes. All timings are in milliseconds.
```{r plot, echo=F}
# timing data
pp = list(m10 = 4.84 / c(4.84, 220.88, 5.99, 5.98, 5.56),
          m100 = 5.32/ c(5.32, 872.21, 6.28, 6.25, 5.84),
          m1000 = 8.39/ c(8.39, 5421.3, 7.91, 7.7, 7.3),
          m10000 = 17.23/ c(17.23, 21675.56, 13.26, 11.45, 10.48))




ppdat = as.data.frame(pp)

xcord = c(500,1000,2500, 5000)
plot(xcord, 
     ppdat[1,], 
     ylim = c(0,5), 
     type = 'l', 
     col = 'red',
     lwd = 2,
     lty = 2,
     xlab = 'No. of rows in Matrix',
     ylab = 'Relative time')


points(xcord, ppdat[2,], type = 'l', col = 'green', lwd = 2)
points(xcord, ppdat[3,], type = 'l', col = 'gray', lwd = 2)
points(xcord, ppdat[4,], type = 'l', col = 'violet', lwd = 2)
points(xcord, ppdat[5,], type = 'l', col = 'black', lwd = 2)


legend('topright', 
       title = 'Functions', 
       legend = c('dist', 'naive_dist',
                  'eucDist', 'r_euc_dist', 'cDist'),
       lwd = 3, 
       col = c('red', 'green','gray', 'violet', 'black'),
       cex = 0.8)


```

# Functions
The function `naive_dist()` is written purely in R using the formula `dist = sqrt((xi-xj)^2 + (yi-yj)^2)`, `dist()` is the base `R` implementation, `dot_dist()` is also written purely R but uses the dot product method of calculating distances, `fastdist2()` is a `C++` implementation of the calculation and `r_euc_dist()`, `eucDist()` and `cDist` are `R` wrapper functions for `C` implementaions of the calculation.

`naive_dist()`
```
naive_dist <- function(x){

#TODO: function documentation

    result = vector('numeric', length = nrow(x) * (nrow(x) - 1)/2)
    k = 1
    for (i in 1:(nrow(x)-1)){
        for (j in (i+1):nrow(x)){
          result[k] = sqrt((x[i,1]-x[j,1])^2 + (x[i,2] - x[j,2])^2)
          k = k + 1
        }
    }
    result
}
```
`dot_dist()`
```
dot_dist <- function(x){
# TODO: function documentation
  result = vector('numeric', length = nrow(x) * (nrow(x) - 1)/2)
  
  i = 1
  m = 1
  while (i <= (nrow(x)-1)){
    for (j in (i+1):nrow(x)){
      k         = x[i,] - x[j,]
      result[m] = sqrt(k %*% k)
      m         = m + 1
    }
    i = i + 1
  }
  result
}
```

`fastdist2()`
```
// filename: fastdist2.cpp

#include <Rcpp.h>
using namespace Rcpp;

//[[Rcpp::export]]

NumericVector fastdist2 (const NumericMatrix &x){
	unsigned int outrows = x.nrow(), i = 0, j = 0, k = 0;

	Rcpp::NumericVector out(outrows * (outrows - 1)/2);

	for (i = 0; i <= outrows - 2; i++){
		for (j = i+1; j <= outrows - 1; j++){

			out(k) = sqrt(sum(pow(x.row(i) - x.row(j), 2.0)));
					k++;
}
	}
	return out;
}
```
`r_euc_dist() and euc_dist()`
```
# filename: r_euc_dist.R
# last edited: 7-AUG-2019


r_euc_dist <- function(matrix, cols_interest = 1:2){

# R wrapper function for euc_dist.c
# check euc_dist.c for source code and function documentation
# input: 
#		matrix: an N by 2 numeric matrix
#		cols_interest : if input is a data.frame, the cols to be used.
		 
# output: a vector of euclidean distances between rows
	cpMat   = matrix[, cols_interest]
	len_mat = dim(cpMat)[1] * dim(cpMat)[2]
    n_rows  = nrow(cpMat)
    vec     = as.vector(t(cpMat))

    .C("euc_dist",
        as.double(vec),
        as.integer(len_mat),
        as.double(vector("double", n_rows * (n_rows - 1)/2)))[[3]]

    }
    
/*
filename: euc_dist.c
last edited: 23-JUL-2019
*/


#include <R.h>
#include <math.h>

/*
Calculates the euclidean distance between rows of a matrix

input: 
    vec: a one dimensional array of length N * 2 where N is the number of rows
       in the input matrix.
    len_vec: N * 2
output:
    result: a one dimensional array of length N * 2 containing the calculated distances 
*/

void euc_dist(double *vec, int *len_vec, double *result)
{
    int k = 0, i, j, iter1 = *len_vec - 3, iter2 = *len_vec - 1;

    for (i = 0; i < iter1; i += 2){
        for (j = i + 2; j < iter2; j += 2){
            result[k] = sqrt((vec[i] - vec[j]) * (vec[i] - vec[j]) 
						+ (vec[i+1] - vec[j+1]) * (vec[i+1] - vec[j+1]));
            k++;
        }
      }

}

```

`eucDist() and eucDist()`
```
# filename: eucDist.R
# wrapper function for eucDist.c
# TODO: error correcting code

eucDist <- function(matrix, cols_interest = 1:2){
	
	cpMat 	= matrix[, cols_interest]
    len_mat = dim(matrix)[1] * dim(matrix)[2]
    n_rows  = nrow(matrix)
    n_cols  = ncol(matrix)
    vec     = as.vector(t(matrix))

    
    .C('eucDist',
       as.double(vec),
       as.integer(n_rows),
       as.integer(n_cols),
       as.double(vector("double", n_rows * (n_rows - 1)/2))
       )[[4]]
}

#include <math.h>
#include <R.h>
#include <Rinternals.h>

/* Euclidean distance */
/*q=.C("eucdist",as.integer(c(1,2,3,4,5,6,7,8)),as.integer(4),as.integer(2),as.double(vector("double",6))*/
void eucDist(double *x, int *m, int *n, double *d)
{
  /* Arguement:
    1.  x is a matrix of dimension n by m
    2.  m is the number of rows
    3.  n is the number of coloums
    4.  d is the pointer for output */
  /*
   d = sqrt(sum((XI-XJ).^2,2));            % Euclidean
   */
  int i,j,k; /* **pointer; /* Indexers */
  int local_m, local_n;
  local_m = *m, local_n = *n;
  double theSum; /* size_t is an unsigned integer of size 16 bits */
  /*
  XI for indexing rows
  XJ for indexing columns
  XI0 unknown for now 
  */
  int  XI, XJ, XI0, index; /* pointers as row indexers*/
  // d = malloc( local_m*(local_m - 1)/2);
  // XI0 =  (double *) x; /* we are not tuching x but using its memory address as XI */
  // x = (double) x;
  index = 0;

  for (i=0; i<local_m-1; ++i) { /* Iterating through the rows of the matrix */
    // XI0 =  XI; /* taking the memory address of the array (Refer to line 29) */
    XI = i*local_n; /* Move along memory by n ( the first coloumn */
    XI0 = XI;
    // Rprintf("XI is %d\n", XI);    
    for (j=i+1; j<local_m; ++j) { /* iterating through the rows from the i_th row*/
      // XI = x + i*(*n); /* Change to XI happpens here after using it on line 28*/
      XJ = j*local_n;
      // Rprintf("XJ is %d\n", XJ);
      // XI = XI0; /* Index? */
      theSum = 0.0;
      for (k=0;k<local_n;k++,++XI,++XJ){
        theSum += pow((x[XI]- x[XJ]), 2.0);
        // Rprintf("x[XI] is %lf and x[XJ] is %lf\n", x[XI], x[XJ]);
        // Rprintf("The sum is %lf\n", theSum);
        // Rprintf("The sum is %d\n", theSum);
      }
      XI = XI0;
      d[index++] = sqrt(theSum);
      // Rprintf("d is %lf\n", d[index]);
      // XI = XI0; /* Index? */
    }
  }
}
```
`cDist() and cDist()`
```
/* 
filename: cDist.c
last edited: 24-JUL-19
*/


/*
 exactly as euc_dist.c but modified to use R's .Call() interface

 input:
    Rvec: a one-dimensional array of length N * 2 where N is the number of rows
          in the input matrix.
    
    resLen: an integer (coerced to double) of the number of pairs
            calculated as N * (N - 1)/2.
 output:
    result: a one-dimensional array containing the calculated distances.  
*/

#include <R.h>
#include <Rinternals.h>
#include <math.h>

SEXP cDist(SEXP Rvec, SEXP resLen){
    int k = 0, i, j, iter1 = length(Rvec) - 3, iter2 = length(Rvec) - 1;

    SEXP result = PROTECT(allocVector(REALSXP, asReal(resLen)));
    double *vec = REAL(Rvec);

    for(i = 0; i < iter1; i += 2){
        for(j = i + 2; j < iter2; j += 2){
            REAL(result)[k] = sqrt(((vec[i] - vec[j]) * (vec[i] - vec[j])) + 
                         ((vec[i+1] - vec[j+1]) * (vec[i+1] - vec[j+1])));
            k++;
        }
    }

    UNPROTECT(1);
    return result;
 
}

# filename: cDist.R
# last edited: 7-AUG-19


# wrapper function for cDist.c
# see cDist.c for function documentation
#
# input: 
#		matrix: an N by 2 numeric matrix
#		cols_interest : if input is a data.frame, the cols to be used.
#
# output: a vector of euclidean distances between the rows of the matrix.
# TODO: error-correcting code



cDist <- function(matrix, cols_interest = 1:2){
	cpMat  = matrix[, cols_interest]
    vec    = as.vector(t(cpMat))
    reslen = nrow(cpMat) * (nrow(cpMat) - 1) * 0.5

    .Call("cDist",
           as.double(vec),
           as.double(reslen))
}
```